---
title: "Diversity analysis at species level"
author: "Jia Xiu (xibeihenai@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---


### Initiate libraries and setup working env
```{r load_packages_01, message=FALSE}

rm(list=ls())

# library(ape)
library(tidyverse)
library(vegan)
library(ggpubr) 
library(phyloseq) 
library(circlize)
library(ggpmisc)  # For regression equation annotation
library(broom)  # For extracting p-values
library(RColorBrewer)


mytheme <- theme_light()+
  theme(text = element_text(size = 10),
        legend.title = element_text(face = "bold", size = 10),
        legend.text = element_text(size = 10), 
        legend.justification = c("right", "top"),
        legend.background = element_rect(),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 10, color = "black"),
        axis.ticks = element_line(colour = "grey70", linewidth = 0.2),
        panel.grid.major = element_line(colour = "grey70", linewidth = 0.1),
        panel.grid.minor = element_blank())  

# Path to the folder containing Kraken2 report files
folder_path <- "div_analysis_species_level/"

# Create the directory if it does not already exist
if (!dir.exists(folder_path)) {
  dir.create(folder_path)
}

```


## 1. alpha-diversity for microcosm samples -  sea & fresh water

Use relative abundance table for the analysis.

```{r}

# read phyloseq object from all the samples - raw reads
ps_rawreads <- readRDS("phyloseq_object_species_rawreads.rds")

# only keep the samples from microcosm experiment
ps_microcosm <- subset_samples(ps_rawreads, category == "microcosm")
# remove taxa do not appear in any samples
ps_microcosm <- filter_taxa(ps_microcosm, function(x) sum(x) != 0, TRUE)


otu_table_all <- as.data.frame(otu_table(ps_rawreads)) 

otu_table <- otu_table_all %>% 
  select(!grep("Mock", names(otu_table_all)) & !grep("control", names(otu_table_all)) & !grep("FieldBlank", names(otu_table_all)) & !grep("f0_b0", names(otu_table_all))) %>% 
  select(!c(IGAPark, MittelmoleWarnemunde, NeptunWerft, SABMarinaBramow, Stadthafen)) %>% 
  rename(f100_b0_f_p0_r1 = Muhlendamm ,
         f0_b100_b_p0_r1 = WarnemundeStrand) 

otu_table[1:2, 1:2]

# number of families x number of samples
dim(otu_table)
# summary of 240 samples (240 microcosm samples and two field samples)
summary(colSums(otu_table))
# total number of reads from 240 samples
sum(colSums(otu_table))

otu_table_rel <- otu_table %>% 
  mutate(across(everything(), ~ .x / sum(.x))) 
otu_table_rel <- t(otu_table_rel)

otu_table_rel[120:121, c(16, 22)]
dim(otu_table_rel)

# get diversity indexes
Shannon <- diversity(otu_table_rel, "shannon")
head(Shannon)
# Simpson <- diversity(t(otu_table_rel), "simpson")
# dim(Simpson)
Richness <- specnumber(otu_table_rel)
head(Richness)
# Pielou's J = H'/ln(S) where H' is Shannon Weiner diversity and S is the total number of species in a sample,
Pielous_evenness <- Shannon/log(Richness)

df <- as.data.frame(cbind(Richness, Shannon)) %>% 
  rownames_to_column() %>%  
  separate(rowname, c("fresh_bac", "sea_bac", "watertype", "passage", "replicate"), "_") %>%  
  select(-replicate) %>% 
  mutate(watertype = factor(watertype, levels = c("f", "b"), labels = c("Freshwater", "Seawater"))) %>% 
  unite(mix, c(fresh_bac, sea_bac), sep = "_", remove = TRUE) %>% 
  mutate(mix = factor(mix, levels = c("f0_b100", "f25_b75", "f50_b50", "f75_b25", "f100_b0"),  labels = c("0:100", "25:75", "50:50", "75:25", "100:0"))) %>% 
  mutate(passage = factor(as.numeric(str_remove(passage, "^p"))))

head(df)

# write.csv(df, paste0(folder_path, "richness_shannon_filtered_all_samples.csv"))

df1 <- df %>% 
  pivot_longer(-c(mix, watertype, passage), names_to = "div_index") %>% 
  group_by(mix, watertype, passage, div_index) %>%
  summarize(mean = mean(value, na.rm=TRUE), sd = round(sd(value), 3), n = round(length(value), 3)) 

df1$se <- df1$sd / sqrt(df1$n)  # Calculate standard error of the mean

head(df1)

(p <- ggplot(df1, aes(passage, mean, group =mix)) +
    geom_errorbar(aes(ymin = mean-se, ymax = mean+se, color = mix), 
                  width = 0.2, position = position_dodge(width = 0.4), alpha = 0.6) +
    geom_line(aes(color = mix), position = position_dodge(width = 0.4), alpha = 0.6) +
    geom_point(aes(fill = mix), size = 2, stroke = 0.1, position = position_dodge(width = 0.4), shape = 21, color = "black") +
    labs(x = "Passages", y = "", title = "") +
    facet_grid(div_index ~ watertype, scales = "free_y") +
    scale_fill_brewer(palette = "RdYlBu", guide=guide_legend(override.aes = list(shape=21)), name = "Mixing ratio \n(fresh:sea)")+
    scale_color_brewer(palette = "RdYlBu", name = "Mixing ratio \n(fresh:sea)")+
    mytheme)

# ggsave(paste0(folder_path, "Fig_S6_richness_shannon_lineplot_species_filtered.png"), width = 5.5, height = 3.5, p)
# ggsave(paste0(folder_path, "Fig_S6_richness_shannon_lineplot_species_filtered.pdf"), width = 5.5, height = 3.5, p)

```


### 2. beta-diversity for only microcosm samples (no passage 0 / field samples)
Here we did PCA analysis based on clr transformed table.
> "Aitchison distance has numerous other desirable properties, such as scale invariance, negating the need to perform rarefaction. This feature is critical when one lacks access to absolute microbial abundance, because scale variant distances ensure equivalence between distances computed from absolute and relative abundance measurements."

```{r}

# read phyloseq object from all the samples - CLR transformed
ps_rclr <- readRDS("phyloseq_object_species_rclr.rds")
ps_rclr

# only keep the samples from microcosm experiment
ps_microcosm <- subset_samples(ps_rclr, category == "microcosm")
# remove taxa do not appear in any samples
ps_microcosm <- filter_taxa(ps_microcosm, function(x) sum(x) != 0, TRUE)

# saveRDS(ps_microcosm, file = "phyloseq_object_species_rclr_microcosm.rds")

# check
ps_microcosm
tail(sample_data(ps_microcosm))

# prepare feature table for further analysis
com <- as.data.frame(otu_table(t(ps_microcosm)))
com[1:3, 1:3]


###########################
# PCA
###########################

# Aitchison distance is simply the Euclidean distance between samples after clr transformation,
pca_result <- rda(com)

sum_pca <- summary(pca_result)
sum_pca$cont$importance[2, 1]
sum_pca$cont$importance[2, 2]

biplot(pca_result, scaling = 2)  # Scaling options: 1 (focus on variables), 2 (focus on sites)

# Site scores (coordinates of samples in PCA space)
df <- as.data.frame(scores(pca_result, display = "sites"))

# Add metadata
df <- df %>% 
  rownames_to_column() %>%  
  separate(rowname, c("fresh_bac", "sea_bac", "watertype", "passage", "replicate"), "_", remove = FALSE) %>%  
  # select(-replicate) %>% 
  mutate(watertype = factor(watertype, levels = c("f", "b"), labels = c("Freshwater", "Seawater"))) %>% 
  unite(mix, c(fresh_bac, sea_bac), sep = "_", remove = TRUE) %>% 
  mutate(mix = factor(mix, levels = c("f0_b100", "f25_b75", "f50_b50", "f75_b25", "f100_b0"),  labels = c("0:100", "25:75", "50:50", "75:25", "100:0"))) %>% 
  mutate(passage = factor(as.numeric(str_remove(passage, "^p")))) # mutate(passage = factor(passage, levels = c("p0", "p1", "p2", "p3", "p4", "p5", "p6"))) 

head(df)

# make plot without arrows

(p <- ggplot(df, aes(PC1, PC2)) +
    geom_point(aes(fill = mix, shape = watertype), size = 3.5, alpha = 0.6, stroke = 0.1) + 
    geom_text(aes(label = passage), size = 2.5, alpha = 0.4, color = "black") +
    labs(x = paste("PCA1 (", round(sum_pca$cont$importance[2, 1] * 100, 2), "%)", sep = ""),
         y = paste("PCA2 (", round(sum_pca$cont$importance[2, 2] * 100, 2), "%)", sep = ""),
         title = "") +
    scale_fill_brewer(palette = "RdYlBu", guide=guide_legend(override.aes = list(shape=21)), name = "Mixing ratio\n(fresh:sea)") +
    scale_shape_manual(values = c(22, 21), name = "Culture water") +
    mytheme)

# ggsave(paste0(folder_path, "Fig_4A_PCA_water_mix_rclr_no_p0_raw.pdf"), width = 4.5, height = 3.2, p)


###########################
# permanova
###########################

# Permutational Multivariate Analysis of Variance Using Distance Matrices

dist <- vegdist(com, method = "euclidean") 

df <- as.data.frame(com[,1:2]) %>% 
  rownames_to_column() %>%  
  select(rowname) %>% 
  separate(rowname, c("fresh_bac", "sea_bac", "watertype", "passage", "replicate"), "_", remove = FALSE) %>%  
  column_to_rownames(var = "rowname") %>% 
  mutate(watertype = factor(watertype, levels = c("f", "b"), labels = c("freshwater", "seawater"))) %>% 
  unite(mix, c(fresh_bac, sea_bac), sep = "_", remove = TRUE) %>% 
  mutate(mix = factor(mix, levels = c("f0_b100", "f25_b75", "f50_b50", "f75_b25", "f100_b0"))) %>% 
  mutate(passage = factor(passage, levels = c("p0", "p1", "p2", "p3", "p4", "p5", "p6")))

head(df)

all(rownames(as.matrix(dist)) == rownames(df))

# # three-way permanova
set.seed(123)
# sequential tests
result_whole <- adonis2(dist ~ mix*watertype*passage, data = df, permutation = 999, by = "terms")
#result_whole <- adonis2(com ~ mix*watertype*passage, data = df, method = "euclidean", permutation = 999)
result_whole

# write.csv(result_whole, paste0(folder_path, "permanova_water_mix_passage_species_Aitchison_rclr.csv"))

```


#### 2.2 Dissimilarity to source communities analysis
To assess the influence of each source community on the coalesced community.

```{r}

# pairwise dissimilarity
dist <- vegdist(com, method = "euclidean", binary = FALSE, diag = 1) 

# Convert distance to dissimilarity in the range [0, 1]
# dissimilarity <- dist / max(dist)

# df <- as.matrix(dissimilarity)
df <- as.matrix(dist)
df <- data.frame(as.table(df))[lower.tri(df, diag = FALSE), ]
cat("For no. of df pairs: Observed = Predict", (240*240-240)/2 == length(df$Freq), "\n")
row.names(df) <- paste(df$Var1, df$Var2, sep = "_")
summary(df)

# define source communities, i.e. f100-f & b100-b 
df_source <- df[grepl("f100_b0_f", df$Var1) | grepl("f100_b0_f", df$Var2) | grepl("f0_b100_b", df$Var1) | grepl("f0_b100_b", df$Var2), ]

df_source <- df_source %>% 
  rownames_to_column(var = "name") %>% 
  separate(name, c("fresh_a", "sea_a", "water_a", "passage_a", "replicate_a", "fresh_b", "sea_b", "water_b", "passage_b", "replicate_b"), "_", remove = FALSE) %>%
  filter(passage_a == passage_b) %>% 
  select(-c(replicate_a, replicate_b, passage_b)) %>%
  rename(passage = passage_a) %>% 
  mutate(passage = factor(gsub("p", "", passage))) %>% 
  unite(a, c(fresh_a, sea_a), sep = "_", remove = TRUE) %>% 
  unite(b, c(fresh_b, sea_b), sep = "_", remove = TRUE) %>% 
  select(-c(Var1, Var2)) %>%
  column_to_rownames(var = "name")

dim(df_source)
tail(df_source)

# in freshwater media
f_levels <- c("f100_b0_f", "f75_b25_f", "f50_b50_f", "f25_b75_f", "f0_b100_f", "f0_b100_b")

df_f <- df_source %>% 
  unite(mix_a, c(a, water_a), sep = "_") %>% 
  unite(mix_b, c(b, water_b), sep = "_") %>% 
  mutate(mix_a = factor(mix_a)) %>% 
  mutate(mix_b = factor(mix_b)) %>% 
  filter(mix_a %in% f_levels & mix_b %in% f_levels) %>% 
  unite(mix, c(mix_a, mix_b), sep = "_") %>% 
  mutate(mix = factor(mix)) %>% 
  filter(mix != "f0_b100_b_f0_b100_b") %>% 
  droplevels()

df_f$watertype <- "Freshwater"

levels(df_f$mix)

# Add source_com and mix_ratio columns
df_f <- df_f %>%
  mutate(
    source_com = ifelse(grepl("f0_b100_b", mix), "sea", "fresh"),
    mix_ratio = case_when(
      grepl("f0_b100_f", mix) ~ "0:100",
      grepl("f25_b75", mix) ~ "25:75",
      grepl("f50_b50", mix) ~ "50:50",
      grepl("f75_b25", mix) ~ "75:25",
      TRUE ~ "100:0"  # Default value
    )
  ) %>% 
  mutate(mix_simple = paste(mix_ratio, "vs.", source_com))

f_factor_levels <- c("100:0 vs. fresh", 
                     "75:25 vs. fresh", 
                     "50:50 vs. fresh", 
                     "25:75 vs. fresh",
                     "0:100 vs. fresh",
                     "100:0 vs. sea",  
                     "75:25 vs. sea", 
                     "50:50 vs. sea", 
                     "25:75 vs. sea", 
                     "0:100 vs. sea")

# I should have 4*4*6 pairs for each coalesced com to a source community, for the source community itself, which should have 6*6 pairs
df_f <- df_f %>% 
  mutate(mix_simple = factor(mix_simple, levels = f_factor_levels)) 

# check results
head(df_f)
dim(df_f) # number of rows should be 900 (96*9 + 36)
df_f %>% filter(mix_ratio == "25:75" & source_com == "sea" & passage == "1") 

# in seawater media
s_levels <- c("f0_b100_b", "f25_b75_b", "f50_b50_b", "f75_b25_b", "f100_b0_b", "f100_b0_f")

df_s <- df_source %>% 
  unite(mix_a, c(a, water_a), sep = "_") %>% 
  unite(mix_b, c(b, water_b), sep = "_") %>% 
  mutate(mix_a = factor(mix_a)) %>% 
  mutate(mix_b = factor(mix_b)) %>% 
  filter(mix_a %in% s_levels & mix_b %in% s_levels) %>% 
  unite(mix, c(mix_a, mix_b), sep = "_") %>% 
  mutate(mix = factor(mix)) %>% 
  filter(mix != "f100_b0_f_f100_b0_f") %>%
  droplevels()

df_s$watertype <- "Seawater"

levels(df_s$mix)

# Add source_com and mix_ratio columns
df_s <- df_s %>%
  mutate(
    source_com = ifelse(grepl("f100_b0_f", mix), "fresh", "sea"),
    mix_ratio = case_when(
      grepl("f100_b0_b", mix) ~ "100:0",
      grepl("f25_b75", mix) ~ "25:75",
      grepl("f50_b50", mix) ~ "50:50",
      grepl("f75_b25", mix) ~ "75:25",
      TRUE ~ "0:100"  # Default value
    )
  ) %>% 
  mutate(mix_simple = paste(mix_ratio, "vs.", source_com))

b_factor_levels <- c("100:0 vs. fresh",
                     "75:25 vs. fresh",
                     "50:50 vs. fresh", 
                     "25:75 vs. fresh", 
                     "0:100 vs. fresh", 
                     "100:0 vs. sea",
                     "75:25 vs. sea",
                     "50:50 vs. sea",
                     "25:75 vs. sea", 
                     "0:100 vs. sea")

df_s <- df_s %>% 
  mutate(mix_simple = factor(mix_simple, levels = b_factor_levels))

# check results
dim(df_s) # number of rows should be 900 (96*9 + 36)
df_s %>% filter(mix_ratio == "0:100" & source_com == "fresh" & passage == "1") 

# combine results from freshwater and seawater
df_fs <- rbind(df_f, df_s) %>% 
  select(mix, watertype, mix_simple, mix_ratio, source_com, passage, Freq)
warning(cat("I should have these much pairs", (9*16+6)*6*2, "\nCheck if the row number is right or not?!\n"))
str(df_fs)



###########################
# box plot 
###########################

# compare two source communities
(p <- ggplot(df_fs, aes(passage, Freq, fill = mix_simple)) +
   geom_boxplot(alpha = 0.8, position = position_dodge(0.8), outlier.size=-1, linewidth = 0.1) +
   geom_point(size = 1, stroke = 0.1, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), shape = 21, color = "black", alpha = 0.6) +
   facet_grid(watertype~.) +
   labs(x = "Passages", y = "Aitchison distance to the native/source communities", title = "") +
   scale_fill_brewer(palette = "RdYlBu", guide=guide_legend(override.aes = list(shape=21)), name = "Mixing ratio vs. native com.\n(fresh:sea)", direction = -1) + 
   mytheme) 
# theme(panel.background = element_rect(fill = "lavenderblush1"), strip.text = element_blank())) 

# ggsave(paste0(folder_path, "distance_to_source_communities_boxplot_rclr_raw1.pdf"), width = 7, height = 3.5, p)


df_fs1 <- df_fs %>% 
  mutate(source_com = factor(source_com, levels = c("fresh", "sea"), labels = c("freshwater community", "seawater community"))) %>% 
  mutate(mix_ratio = factor(mix_ratio, levels = c("100:0", "75:25", "50:50", "25:75", "0:100" )))

# compare two culturing water conditions
(p <- ggplot(df_fs1, aes(interaction(watertype, passage), Freq, fill = mix_ratio)) +
    geom_boxplot(alpha = 0.8, position = position_dodge(0.8), outlier.size=-1, linewidth = 0.1) +
    geom_point(size = 1, stroke = 0.1, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), shape = 21, color = "black", alpha = 0.6) +
    facet_grid(source_com~.) +
    labs(x = "Culturing water x Passages", y = "Aitchison distance to the native communities", title = "") +
    scale_fill_brewer(palette = "RdYlBu", guide=guide_legend(override.aes = list(shape=21)), name = "Mixing ratio vs. native com.\n(fresh:sea)", direction = -1) + 
    mytheme ) # + theme(axis.text.x = element_text(angle = 45, hjust = 1)))

# ggsave(paste0(folder_path, "distance_to_source_communities_boxplot_rclr_raw2.pdf"), width = 9, height = 3.8, p)


###########################
# line plot for all passages
###########################

df_fs1 <- df_fs %>% 
  separate(mix_ratio, c("fresh_ratio", "sea_ratio"), ":", remove = FALSE) %>% 
  select(-c(mix, mix_simple))

df_fs1 <- df_fs1 %>%
  mutate(source_ratio = if_else(source_com == "fresh", fresh_ratio, sea_ratio)) 

df_fs1$source_ratio <- as.numeric(as.character(df_fs1$source_ratio))
df_fs1$fresh_ratio <- as.numeric(as.character(df_fs1$fresh_ratio))

head(df_fs1)

df_fs_lm <- df_fs1 %>% 
  select(-c(fresh_ratio, sea_ratio)) %>% 
  mutate(across(where(~ !is.numeric(.x)), as.factor)) 

# write down the dataframe to have a verification
# write.csv(df_fs_lm, paste0(folder_path, "Aitchison_distance_to_source_coms_rclr_raw.csv"))
str(df_fs_lm)
df_fs_lm[c(1, 10, 40,  1031, 1001, 1010, 1040, 1060, 1700), ]

# linear regressions (Freq ~ source_ratio) for each combination of watertype and source_com
results <- df_fs_lm %>%
  group_by(watertype, source_com) %>%
  do(tidy(lm(Freq ~ source_ratio, data = .)))

results

# check r.squared for each watertype × source_com
df_fs_lm %>%
  group_by(watertype, source_com) %>%
  do(glance(lm(Freq ~ source_ratio, data = .))) %>%
  select(watertype, source_com, r.squared, adj.r.squared)

#########################################
# # Step 1: Compute p-values for each group
# lm_results <- df_fs1 %>%
#   group_by(watertype, source_com) %>%
#   do({
#     model <- lm(Freq ~ source_ratio, data = .)
#     tidy(model)  # Extract regression statistics
#   }) %>%
#   filter(term == "source_ratio") %>%  # Only keep slope term (not intercept)
#   mutate(significant = p.value < 0.05)  # Mark significant regressions
# 
# # Step 2: Merge with original data to keep significant groups only
# df_filtered <- df_fs1 %>%
#   inner_join(lm_results, by = c("watertype", "source_com")) %>% 
#   mutate(source_com = factor(source_com, levels = c("fresh", "sea"), labels = c("Freshwater", "Seawater")))
# head(df_filtered)

df_fs1 <- df_fs1 %>% 
  mutate(source_com = factor(source_com, levels = c("fresh", "sea"), labels = c("Freshwater", "Seawater")))
str(df_fs1)

# Step 3: line plot 

# line plot - combine all passages
# update the regression results by the correct one above
(p <- ggplot(df_fs1, aes(fresh_ratio, Freq, fill = source_com, color = source_com)) +
    geom_point(size = 1.5, alpha = 0.05, position = position_jitter(width = 4)) +
    scale_x_continuous( breaks = c(0, 25, 50, 75, 100),  labels = c("0:100", "25:75", "50:50", "75:25", "100:0")) +
    facet_grid(. ~ watertype ) +
    geom_smooth(data = df_fs1, method = "lm", aes(color = source_com), size = 0.5, alpha = 0.3) +  
    # geom_smooth(method = "lm", aes(color = source_com), alpha = 0.6) +  # Match regression line colors
    # Plot regression lines only for significant groups
    # Add regression equation, R², and p-value
    stat_poly_eq(
      aes(label = after_stat(paste(..eq.label.., ..rr.label.., ..p.value.label.., sep = "~~~"))),
      formula = y ~ x, 
      parse = TRUE, 
      size = 2.5) +
    labs(x = "Mixing ratio\n(fresh:sea)", y = "Aitchison distance", title = "") +
    scale_fill_manual(values = c("#4575B4", "#D73027"), guide=guide_legend(override.aes = list(shape=21)), name = "Distance to") +
    scale_color_manual(values = c("#4575B4", "#D73027"), name = "Distance to") +
    mytheme +
    theme(legend.position.inside = c(0.95, 0.95)))

# ggsave(paste0(folder_path, "Fig_4C_distance_to_source_coms_line_plot_rclr_raw.pdf"), width = 5.8, height = 3, p)




###########################
# dis to source on two axis
###########################

# Summarize mean and standard error for fresh and sea
df_summary <- df_fs %>%
  select(-c(mix, mix_simple)) %>% 
  group_by(watertype, mix_ratio, passage, source_com) %>%  # Group by relevant factors
  summarise(
    mean = mean(Freq, na.rm = TRUE),
    se = sd(Freq, na.rm = TRUE) / sqrt(n())) %>%
  ungroup() %>% 
  arrange(passage, mix_ratio) %>% 
  pivot_wider(names_from = source_com, values_from = c(mean, se), names_glue = "{.value}_{source_com}") %>% 
  mutate(mix_ratio = factor(mix_ratio, levels = c("0:100", "25:75", "50:50", "75:25", "100:0")))

summary(df_summary)
# y = -x+a
a <- max(df_summary$mean_fresh) + min(df_summary$mean_sea)

min_xylim <- min(min(df_summary$mean_sea), min(df_summary$mean_fresh))
max_xylim <- max(max(df_summary$mean_sea), max(df_summary$mean_fresh))

# extract the color codes from the "RdYlBu" palette using the RColorBrewer
# display.brewer.pal(5, "RdYlBu")
# (colors <- brewer.pal(5, "RdYlBu"))
colors <- c("#D7191C", "#FDAE61", "gold2", "#ABD9E9", "#2C7BB6")

# Plot mean and standard error with connections
(p <- ggplot(df_summary, aes(x = mean_fresh, y = mean_sea, shape = passage)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "darkred", alpha = 0.2, size = 0.4) +  # y = x line
    # geom_abline(slope = -1, intercept = a, linetype = "dashed", color = "darkblue", alpha = 0.2, size = 0.4) +  # y = x line
    geom_errorbar(aes(ymin = mean_sea - se_sea, ymax = mean_sea + se_sea, color = mix_ratio), linewidth = 0.2, width = 0.1, alpha = 0.1, position = position_dodge(0.3)) +  # SE as error bars
    geom_errorbarh(aes(xmin = mean_fresh - se_fresh, xmax = mean_fresh + se_fresh, color = mix_ratio), linewidth = 0.2, height = 0.1, alpha = 0.1, position = position_dodge(0.3)) +  # Horizontal SE error bars
    geom_path(aes(group = mix_ratio, color = mix_ratio), linewidth = 0.2, alpha= 0.2) +
    # geom_point(aes(fill = mix_ratio, shape = watertype), size = 1, stroke = 0.1, alpha= 0.7, color = "black", shape = 21) +
    geom_text(aes(label = passage, color = mix_ratio), size = 3) +
    facet_grid(. ~ watertype) +
    labs(x = "Aitchison to freshwater source communities", y = "Aitchison to seawater source communities", title = "") +
    scale_color_manual(values = colors, name = "Mixing ratio \n(fresh:sea)") +
    xlim(min_xylim, max_xylim) +  
    ylim(min_xylim, max_xylim) +  
    mytheme)

# ggsave(paste0(folder_path, "Fig_4B_distance_to_source_coms_mean_se_rclr_raw.pdf"), plot = p, width = 5.5, height = 2.8)

```


### Session Info
```{r}
sessionInfo()
```