---
title: "Feature table preparation as phyloseq object"
author: "Jia Xiu (xibeihenai@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

> I used the emu for taxonomic annotation (at species level) for the microbiome of Coalescence experiment (16S full-length rRNA gene sequencing by Nanopore).

## 1. Initiate libraries
```{r load_packages_01, message=FALSE}

rm(list=ls())

library(tidyverse)
library(phyloseq) # for prepare phyloseq object for network analysis
library(compositions)  # for CLR

```


## 2. Load metadata file
```{r}

# load metadata file of all samples
metadata <- read.csv("metadata_16S_17-08-2024.tsv", header = 1, sep = "\t")
metadata <- metadata %>% 
  select(-c(X27FBarcodeNumber, X1492RBarcodeNumber)) %>% 
  select(-sample.id) %>% 
  column_to_rownames(var = "SampleName")

tail(metadata)

```


## 3. load raw ouput from EMU

```{r}
# check how many reads annotated as bacteria
raw_species_table <- read.table("combined_estimated_counts.tsv", header = TRUE, sep = "\t") 

dim(raw_species_table)
raw_species_table[1:3, 1:9]

taxa_table <- raw_species_table %>% 
  select(species, genus, family, order, class, phylum) %>% 
  column_to_rownames(var = "species")

tail(taxa_table)
dim(taxa_table)

```

## 4. Species level
### 4.1 Community table with raw reads 
```{r}
species_table <- raw_species_table %>% 
  select(-c(tax_id, genus, family, order, class, phylum)) %>% 
  column_to_rownames(var = "species")

# write.csv(colSums(species_table), "total_number_of_reads_species_level.csv")

# change the estimate reads as integer
rownames_saved <- rownames(species_table)
species_table <- as.data.frame(apply(species_table, 2, as.integer))
rownames(species_table) <- rownames_saved

species_table <- species_table[rownames(species_table) != "unassigned", ]

dim(species_table)
species_table[1:3, 1:4]

# Make phyloseq object - raw reads
otu_table_phyloseq <- otu_table(as.matrix(species_table), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table))

ps_rawreads <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rawreads

# saveRDS(ps_rawreads, file = "phyloseq_object_species_rawreads.rds")


# Clean the feature table by removing the low abundant species 
# Define the threshold value to remove species with a total abundance less than 3 and that occur in fewer than 3 samples,
threshold <- 3

# Remove rows where the sum and occurence of all columns is less than the threshold
species_table_filtered <- species_table %>%
  filter(rowSums(across(everything())) >= threshold) %>% 
  filter(rowSums(across(everything(), ~ . > 0)) >= threshold)

species_table_filtered[1:2, 1:2]
dim(species_table_filtered)

# Total number of reads per sample
summary(colSums(species_table_filtered))

# Total reads per family
summary(rowSums(species_table_filtered))


taxa_table_filtered <- taxa_table[rownames(taxa_table) %in% rownames(species_table_filtered), ]
dim(taxa_table_filtered)


# Make phyloseq object - raw reads
# non-filtered
otu_table_phyloseq <- otu_table(as.matrix(species_table), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table))

ps_rawreads <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rawreads

# saveRDS(ps_rawreads, file = "phyloseq_object_species_rawreads.rds")

# filtered
otu_table_phyloseq <- otu_table(as.matrix(species_table_filtered), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_filtered))

ps_rawreads_filtered <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rawreads_filtered

# saveRDS(ps_rawreads_filtered, file = "phyloseq_object_species_filtered_rawreads.rds")

```
Before filtering, there are `r nrow(species_table)` species. After filtering species with a total abundance less than 3 and that occur in fewer than 3 samples, we got `r nrow(species_table_filtered)` species.



### 4.2 CLR

#### 4.2.1 CLR transformation on filtered table
**compositional data** everything that is part of a whole is compositional. the change in one component will affect the other component. By sequencing, you are subsampling. 
CLR normalization transforms compositional data to address the "closed nature" of relative abundances, $$CLR(xi)=ln(xi/g(x))$$

1. CLR produces values on a continuous scale that are suitable for Euclidean geometry, making it compatible with many statistical methods (e.g., PCA). 
2. CLR handles compositional bias effectively but can be sensitive to zeros since the logarithm is undefined for zero. 
3. The sum of each sample in CLR (Centered Log-Ratio) and rCLR (Robust Centered Log-Ratio) transformations will always be zero.

**Robust CLR Transformation**\
The Robust CLR Transformation is a modification of the standard CLR transformation designed to handle issues that may arise when some components of the compositional data have values of zero or near-zero. Zero values in compositional data can cause the CLR transformation to be undefined because the logarithm of zero is not defined.\
To avoid issues with zeros, the Robust CLR typically uses a small constant (e.g., adding a small pseudo-count) or applies a more robust estimation method for the geometric mean, such as ignoring zeros or using a trimmed mean. This approach makes the CLR transformation more stable and applicable in real-world situations where zero values are common in compositional data (e.g., microbiome studies, where some taxa may be absent in certain samples).\

**Steps in Robust CLR:**\
-   1. Replace zeros with a small constant: One common method is to replace zeros with a small positive value, such as the smallest observed value in the dataset or a predefined constant (like $e= 1e-6$).
-   2. Modified geometric mean: In the robust CLR, the geometric mean is computed on the adjusted data (with small constants added to avoid zeros), ensuring stability during the transformation.
$$CLR(xi)=ln((xi + e)/g(x + e))$$.

```{r}
####################################
# Option 1. CLR transformation
# Robust CLR (rCLR): The rCLR transformation modifies the CLR approach by adding a small constant (often called a pseudocount) to the data before performing the log transformation, making it robust to zeros. This approach is especially useful in microbiome data where zeros are common in OTU tables.We performed a centered log-ratio (CLR) transformation of feature table, which has been added a small pseudocount, 10^-6. 
otu_table <- as.data.frame(t(species_table_filtered)) # samples in rows, species in columns

# Add pseudocount to handle zeros
pseudocount <- 1e-6
otu_table <- otu_table + pseudocount

# CLR Transformation
species_table_rclr <- t(apply(otu_table, 1, function(x) log(x) - mean(log(x))))
species_table_rclr <- as.data.frame(t(species_table_rclr))

# check if the sum of each sample approximate zero
range(round(colSums(species_table_rclr), 5))

species_table_rclr[1:2, 1]

# Make phyloseq object - rclr
otu_table_phyloseq <- otu_table(as.matrix(species_table_rclr), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_filtered))

ps_rclr <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rclr

# saveRDS(ps_rclr, file = "phyloseq_object_species_filtered_rclr.rds")

####################################
# Option 2 CLR # based on https://www.youtube.com/watch?v=ulo7WatBEAo&ab_channel=RiffomonasProject
# The CLR transformation does not replace zeros. Instead, only non-zero features are transformed, using the geometric mean of non-zero features as denominator.

# geometric mean for only non-zero values
gm <- function(x) {
  exp(mean(log(x[x>0])))
}


# robust clr remove zeros
species_table_clr <- species_table_filtered %>%
  as.data.frame() %>%
  rownames_to_column(var = "Family") %>%
  pivot_longer(-Family, values_to = "reads", names_to = "Sample") %>%
  filter(reads != 0) %>%
  group_by(Sample) %>%
  mutate(rclr = log(reads/gm(reads))) %>%
  ungroup() %>%
  select(-reads) %>%
  pivot_wider(names_from = "Sample", values_from = "rclr", values_fill = 0) %>% # if there are any NA values, we make them zeros
  as.data.frame() %>%
  column_to_rownames(var = "Family")

species_table_clr[1:2, 1]
range(round(colSums(species_table_clr), 5))

# Make phyloseq object - clr
otu_table_phyloseq <- otu_table(as.matrix(species_table_clr), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_filtered))

ps_clr <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_clr

# saveRDS(ps_clr, file = "phyloseq_object_species_filtered_clr.rds")

####################################
# Option 3 CLR-transformation using r-package compositions

library(compositions)  # for CLR


# Assume you already have a phyloseq object
# For example:
# ps <- readRDS("your_phyloseq_object.rds")

# Step 1: Extract the abundance table (OTU/ASV)
otu <- otu_table(ps_rawreads_filtered)
otu_mat <- as(otu, "matrix")

# Step 2: Transpose if taxa are columns
if (taxa_are_rows(ps_rawreads_filtered)) {
  otu_mat <- t(otu_mat)
}

# Step 3: Add a pseudocount to avoid zeros
otu_pseudo <- otu_mat + pseudocount

# Step 4: CLR transform (compositions package works with data frames)
otu_clr <- compositions::clr(otu_pseudo)

# Step 5: Create phyloseq-compatible OTU table. Transpose back if needed and wrap in otu_table
otu_clr_table <- otu_table(t(otu_clr), taxa_are_rows = TRUE)


# Check that taxon names match
all(rownames(tax_table(ps_rawreads_filtered)) %in% rownames(otu_clr_table))

# If this returns FALSE, inspect:
head(rownames(tax_table(ps_rawreads_filtered)))
head(rownames(otu_clr_table))


# Step 6 (Optional): Create a new phyloseq object with CLR-transformed data
ps_clr <- phyloseq(otu_clr_table, tax_table(ps_rawreads_filtered), sample_data(ps_rawreads_filtered))

# Now ps_clr contains CLR-transformed abundances
# Compare what I did manually

# raw table
otu[7:25, c(1,7)]

# what I did manually
ps_rclr <- readRDS("phyloseq_object_species_filtered_rclr.rds")
otu_rclr <- otu_table(ps_rclr)
otu_rclr[7:25, c(1,7)]

#from the package
otu_clr_table[7:25, c(1,7)]
```

#### 4.2.2. CLR transformation on non-filtered table
```{r}
##############################
# Option 1. Robust CLR (rCLR)
##############################
# The rCLR transformation modifies the CLR approach by adding a small constant (often called a pseudocount) to the data before performing the log transformation, making it robust to zeros. This approach is especially useful in microbiome data where zeros are common in OTU tables.We performed a centered log-ratio (CLR) transformation of feature table, which has been added a small pseudocount, 10^-6. 
otu_table <- as.data.frame(t(species_table)) # samples in rows, species in columns

# Add pseudocount to handle zeros
pseudocount <- 1e-6
otu_table <- otu_table + pseudocount

# CLR Transformation
species_table_rclr <- t(apply(otu_table, 1, function(x) log(x) - mean(log(x))))
species_table_rclr <- as.data.frame(t(species_table_rclr))

# check if the sum of each sample approximate zero
range(round(colSums(species_table_rclr), 5))

# species_table_rclr[1:2, 1]

# Make phyloseq object - rclr
otu_table_phyloseq <- otu_table(as.matrix(species_table_rclr), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table))

ps_rclr <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rclr

# saveRDS(ps_rclr, file = "phyloseq_object_species_rclr.rds")

###################################
# Option 2. The CLR transformation
###################################
# based on https://www.youtube.com/watch?v=ulo7WatBEAo&ab_channel=RiffomonasProject
# The CLR transformation does not replace zeros. Instead, only non-zero features are transformed, using the geometric mean of non-zero features as denominator.

# geometric mean for only non-zero values
gm <- function(x) {
  exp(mean(log(x[x>0])))
}


# robust clr remove zeros
species_table_clr <- species_table %>%
  as.data.frame() %>%
  rownames_to_column(var = "Family") %>%
  pivot_longer(-Family, values_to = "reads", names_to = "Sample") %>%
  filter(reads != 0) %>%
  group_by(Sample) %>%
  mutate(rclr = log(reads/gm(reads))) %>%
  ungroup() %>%
  select(-reads) %>%
  pivot_wider(names_from = "Sample", values_from = "rclr", values_fill = 0) %>% # if there are any NA values, we make them zeros
  as.data.frame() %>%
  column_to_rownames(var = "Family")

species_table_clr[1:2, 1]
range(round(colSums(species_table_clr), 5))

# Make phyloseq object - clr
otu_table_phyloseq <- otu_table(as.matrix(species_table_clr), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table))

ps_clr <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_clr

# saveRDS(ps_clr, file = "phyloseq_object_species_clr.rds")

####################################
# Option 3 CLR-transformation using r-package compositions
####################################

# Assume you already have a phyloseq object
# For example:
# ps <- readRDS("your_phyloseq_object.rds")

# Step 1: Extract the abundance table (OTU/ASV)
otu <- otu_table(ps_rawreads)
otu_mat <- as(otu, "matrix")

# Step 2: Transpose if taxa are columns
if (taxa_are_rows(ps_rawreads)) {
  otu_mat <- t(otu_mat)
}

# Step 3: Add a pseudocount to avoid zeros
otu_pseudo <- otu_mat + pseudocount

# Step 4: CLR transform (compositions package works with data frames)
otu_clr <- compositions::clr(otu_pseudo)

# Step 5: Create phyloseq-compatible OTU table. Transpose back if needed and wrap in otu_table
otu_clr_table <- otu_table(t(otu_clr), taxa_are_rows = TRUE)

# Check that taxon names match
all(rownames(tax_table(ps_rawreads)) %in% rownames(otu_clr_table))

# If this returns FALSE, inspect:
head(rownames(tax_table(ps_rawreads)))
head(rownames(otu_clr_table))

# Step 6 (Optional): Create a new phyloseq object with CLR-transformed data
ps_clr <- phyloseq(otu_clr_table, tax_table(ps_rawreads), sample_data(ps_rawreads))

# Now ps_clr contains CLR-transformed abundances
# Compare what I did manually

# raw table
otu[c(1:7, 50:55), c(1,7)]

# what I did manually
ps_rclr <- readRDS("phyloseq_object_species_rclr.rds")
otu_rclr <- otu_table(ps_rclr)
otu_rclr[c(1:7, 50:55), c(1,7)]

#from the package
otu_clr_table[c(1:7, 50:55), c(1,7)]
```

### 4.3 Relative abundance 
```{r}

# read tsv table
species_table <- read.table("combined_rel_abundance.tsv", header = TRUE, sep = "\t") 

dim(species_table)
species_table[1:3, 1:9]

taxa_table <- species_table %>% 
  select(species, genus, family, order, class, phylum) %>% 
  column_to_rownames(var = "species")

tail(taxa_table)
dim(taxa_table)

species_table <- species_table %>% 
  select(-c(tax_id, genus, family, order, class, phylum)) %>% 
  column_to_rownames(var = "species")

dim(species_table)
species_table[1:3, 1:4]

species_table <- species_table[rownames(species_table) != "unassigned", ]

species_table <- species_table * 100

# check if all the species >0 
species_table_filtered <- species_table %>%
  filter(rowSums(across(everything())) >= 0) 
dim(species_table_filtered)

# Total number of reads per sample
summary(colSums(species_table))

# Total reads per family
summary(rowSums(species_table))


taxa_table_filtered <- taxa_table[rownames(taxa_table) %in% rownames(species_table_filtered), ]
dim(taxa_table_filtered)


# Make phyloseq object - raw reads
otu_table_phyloseq <- otu_table(as.matrix(species_table), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table))

ps_rel <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rel

# saveRDS(ps_rel, file = "phyloseq_object_species_rel.rds")

```


## 5. Genus level
### 5.1 Community table with raw reads 
```{r}

# taxonomy table at genus level
taxa_table_genus <- taxa_table[!duplicated(taxa_table$genus), ] %>% 
  mutate(genus = ifelse(is.na(genus), "Unclassified", genus)) %>% 
  filter(! genus == "Unclassified") %>% 
  rownames_to_column(var = "species") %>% 
  select(-species) %>% 
  column_to_rownames(var = "genus")

dim(taxa_table_genus)
head(taxa_table_genus)


# feature table
genus_table <- raw_species_table %>% 
  column_to_rownames(var = "tax_id") %>% 
  select(-c(species, family, order, class, phylum)) %>%
  mutate(genus = ifelse(is.na(genus), "Unclassified", genus)) %>%  # <-- fill NA
  group_by(genus) %>%
  summarise(across(where(is.numeric), sum)) %>% 
  as.data.frame() %>% 
  column_to_rownames(var = "genus") 

# write.csv(colSums(genus_table), "total_number_of_reads_genus_level.csv")

# change the estimate reads as integer
rownames_saved <- rownames(genus_table)
genus_table <- as.data.frame(apply(genus_table, 2, as.integer))
rownames(genus_table) <- rownames_saved

genus_table <- genus_table[rownames(genus_table) != "unassigned", ]

dim(genus_table)
genus_table[1:3, 1:4]

# Make phyloseq object - raw reads
# non-filtered
otu_table_phyloseq <- otu_table(as.matrix(genus_table), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_genus))

ps_rawreads <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rawreads

# saveRDS(ps_rawreads, file = "phyloseq_object_genus_rawreads.rds")


# Clean the feature table by removing the low abundant genus 
# Define the threshold value to remove genus with a total abundance less than 3 and that occur in fewer than 3 samples,
threshold <- 3

# Remove rows where the sum and occurence of all columns is less than the threshold
genus_table_filtered <- genus_table %>%
  filter(rowSums(across(everything())) >= threshold) %>% 
  filter(rowSums(across(everything(), ~ . > 0)) >= threshold)

genus_table_filtered[1:2, 1:2]
dim(genus_table_filtered)

# Total number of reads per sample
summary(colSums(genus_table_filtered))

# Total reads per family
summary(rowSums(genus_table_filtered))


taxa_table_genus_filtered <- taxa_table_genus[rownames(taxa_table_genus) %in% rownames(genus_table_filtered), ]
dim(taxa_table_genus_filtered)


# Make phyloseq object - raw reads
# filtered
otu_table_phyloseq <- otu_table(as.matrix(genus_table_filtered), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_genus_filtered))

ps_rawreads_filtered <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rawreads_filtered

# saveRDS(ps_rawreads_filtered, file = "phyloseq_object_genus_filtered_rawreads.rds")

```
Before filtering, there are `r nrow(genus_table)` genus. After filtering genus with a total abundance less than 3 and that occur in fewer than 3 samples, we got `r nrow(genus_table_filtered)` genus.



### 5.2 CLR

#### 5.2.1 CLR transformation on filtered table

```{r}
####################################
# Option 1. CLR transformation
# Robust CLR (rCLR): The rCLR transformation modifies the CLR approach by adding a small constant (often called a pseudocount) to the data before performing the log transformation, making it robust to zeros. This approach is especially useful in microbiome data where zeros are common in OTU tables.We performed a centered log-ratio (CLR) transformation of feature table, which has been added a small pseudocount, 10^-6. 
otu_table <- as.data.frame(t(genus_table_filtered)) # samples in rows, genus in columns

# Add pseudocount to handle zeros
pseudocount <- 1e-6
otu_table <- otu_table + pseudocount

# CLR Transformation
genus_table_rclr <- t(apply(otu_table, 1, function(x) log(x) - mean(log(x))))
genus_table_rclr <- as.data.frame(t(genus_table_rclr))

# check if the sum of each sample approximate zero
range(round(colSums(genus_table_rclr), 5))

genus_table_rclr[1:2, 1]

# Make phyloseq object - rclr
otu_table_phyloseq <- otu_table(as.matrix(genus_table_rclr), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_genus_filtered))

ps_rclr <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rclr

# saveRDS(ps_rclr, file = "phyloseq_object_genus_filtered_rclr.rds")

####################################
# Option 2 CLR # based on https://www.youtube.com/watch?v=ulo7WatBEAo&ab_channel=RiffomonasProject
# The CLR transformation does not replace zeros. Instead, only non-zero features are transformed, using the geometric mean of non-zero features as denominator.

# geometric mean for only non-zero values
gm <- function(x) {
  exp(mean(log(x[x>0])))
}


# robust clr remove zeros
genus_table_clr <- genus_table_filtered %>%
  as.data.frame() %>%
  rownames_to_column(var = "Family") %>%
  pivot_longer(-Family, values_to = "reads", names_to = "Sample") %>%
  filter(reads != 0) %>%
  group_by(Sample) %>%
  mutate(rclr = log(reads/gm(reads))) %>%
  ungroup() %>%
  select(-reads) %>%
  pivot_wider(names_from = "Sample", values_from = "rclr", values_fill = 0) %>% # if there are any NA values, we make them zeros
  as.data.frame() %>%
  column_to_rownames(var = "Family")

genus_table_clr[1:2, 1]
range(round(colSums(genus_table_clr), 5))

# Make phyloseq object - clr
otu_table_phyloseq <- otu_table(as.matrix(genus_table_clr), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_genus_filtered))

ps_clr <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_clr

# saveRDS(ps_clr, file = "phyloseq_object_genus_filtered_clr.rds")

####################################
# Option 3 CLR-transformation using r-package compositions

library(compositions)  # for CLR


# Assume you already have a phyloseq object
# For example:
# ps <- readRDS("your_phyloseq_object.rds")

# Step 1: Extract the abundance table (OTU/ASV)
otu <- otu_table(ps_rawreads_filtered)
otu_mat <- as(otu, "matrix")

# Step 2: Transpose if taxa are columns
if (taxa_are_rows(ps_rawreads_filtered)) {
  otu_mat <- t(otu_mat)
}

# Step 3: Add a pseudocount to avoid zeros
otu_pseudo <- otu_mat + pseudocount

# Step 4: CLR transform (compositions package works with data frames)
otu_clr <- compositions::clr(otu_pseudo)

# Step 5: Create phyloseq-compatible OTU table. Transpose back if needed and wrap in otu_table
otu_clr_table <- otu_table(t(otu_clr), taxa_are_rows = TRUE)


# Check that taxon names match
all(rownames(tax_table(ps_rawreads_filtered)) %in% rownames(otu_clr_table))

# If this returns FALSE, inspect:
head(rownames(tax_table(ps_rawreads_filtered)))
head(rownames(otu_clr_table))


# Step 6 (Optional): Create a new phyloseq object with CLR-transformed data
ps_clr <- phyloseq(otu_clr_table, tax_table(ps_rawreads_filtered), sample_data(ps_rawreads_filtered))

# Now ps_clr contains CLR-transformed abundances
# Compare what I did manually

# raw table
otu[7:25, c(1,7)]

# what I did manually
ps_rclr <- readRDS("phyloseq_object_genus_filtered_rclr.rds")
otu_rclr <- otu_table(ps_rclr)
otu_rclr[7:25, c(1,7)]

#from the package
otu_clr_table[7:25, c(1,7)]

```

#### 5.2.2. CLR transformation on non-filtered table
```{r}
##############################
# Option 1. Robust CLR (rCLR)
##############################
# The rCLR transformation modifies the CLR approach by adding a small constant (often called a pseudocount) to the data before performing the log transformation, making it robust to zeros. This approach is especially useful in microbiome data where zeros are common in OTU tables.We performed a centered log-ratio (CLR) transformation of feature table, which has been added a small pseudocount, 10^-6. 
otu_table <- as.data.frame(t(genus_table)) # samples in rows, genus in columns

# Add pseudocount to handle zeros
pseudocount <- 1e-6
otu_table <- otu_table + pseudocount

# CLR Transformation
genus_table_rclr <- t(apply(otu_table, 1, function(x) log(x) - mean(log(x))))
genus_table_rclr <- as.data.frame(t(genus_table_rclr))

# check if the sum of each sample approximate zero
range(round(colSums(genus_table_rclr), 5))

# genus_table_rclr[1:2, 1]

# Make phyloseq object - rclr
otu_table_phyloseq <- otu_table(as.matrix(genus_table_rclr), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_genus))

ps_rclr <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rclr

# saveRDS(ps_rclr, file = "phyloseq_object_genus_rclr.rds")

###################################
# Option 2. The CLR transformation
###################################
# based on https://www.youtube.com/watch?v=ulo7WatBEAo&ab_channel=RiffomonasProject
# The CLR transformation does not replace zeros. Instead, only non-zero features are transformed, using the geometric mean of non-zero features as denominator.

# geometric mean for only non-zero values
gm <- function(x) {
  exp(mean(log(x[x>0])))
}


# robust clr remove zeros
genus_table_clr <- genus_table %>%
  as.data.frame() %>%
  rownames_to_column(var = "Family") %>%
  pivot_longer(-Family, values_to = "reads", names_to = "Sample") %>%
  filter(reads != 0) %>%
  group_by(Sample) %>%
  mutate(rclr = log(reads/gm(reads))) %>%
  ungroup() %>%
  select(-reads) %>%
  pivot_wider(names_from = "Sample", values_from = "rclr", values_fill = 0) %>% # if there are any NA values, we make them zeros
  as.data.frame() %>%
  column_to_rownames(var = "Family")

genus_table_clr[1:2, 1]
range(round(colSums(genus_table_clr), 5))

# Make phyloseq object - clr
otu_table_phyloseq <- otu_table(as.matrix(genus_table_clr), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_genus))

ps_clr <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_clr

# saveRDS(ps_clr, file = "phyloseq_object_genus_clr.rds")

####################################
# Option 3 CLR-transformation using r-package compositions
####################################

# Assume you already have a phyloseq object
# For example:
# ps <- readRDS("your_phyloseq_object.rds")

# Step 1: Extract the abundance table (OTU/ASV)
otu <- otu_table(ps_rawreads)
otu_mat <- as(otu, "matrix")

# Step 2: Transpose if taxa are columns
if (taxa_are_rows(ps_rawreads)) {
  otu_mat <- t(otu_mat)
}

# Step 3: Add a pseudocount to avoid zeros
otu_pseudo <- otu_mat + pseudocount

# Step 4: CLR transform (compositions package works with data frames)
otu_clr <- compositions::clr(otu_pseudo)

# Step 5: Create phyloseq-compatible OTU table. Transpose back if needed and wrap in otu_table
otu_clr_table <- otu_table(t(otu_clr), taxa_are_rows = TRUE)

# Check that taxon names match
all(rownames(tax_table(ps_rawreads)) %in% rownames(otu_clr_table))

# If this returns FALSE, inspect:
head(rownames(tax_table(ps_rawreads)))
head(rownames(otu_clr_table))

# Step 6 (Optional): Create a new phyloseq object with CLR-transformed data
ps_clr <- phyloseq(otu_clr_table, tax_table(ps_rawreads), sample_data(ps_rawreads))

# Now ps_clr contains CLR-transformed abundances
# Compare what I did manually

# raw table
otu[c(1:7, 50:55), c(1,7)]

# what I did manually
ps_rclr <- readRDS("phyloseq_object_genus_rclr.rds")
otu_rclr <- otu_table(ps_rclr)
otu_rclr[c(1:7, 50:55), c(1,7)]

#from the package
otu_clr_table[c(1:7, 50:55), c(1,7)]
```

### 4.3 Relative abundance 
```{r}

# read tsv table
genus_table <- read.table("combined_rel_abundance.tsv", header = TRUE, sep = "\t") 

dim(genus_table)
genus_table[1:3, 1:9]


genus_table <- genus_table %>% 
  select(-c(tax_id, genus, family, order, class, phylum)) %>% 
  column_to_rownames(var = "genus")

dim(genus_table)
genus_table[1:3, 1:4]

genus_table <- genus_table[rownames(genus_table) != "unassigned", ]

genus_table <- genus_table * 100

# check if all the genus >0 
genus_table_filtered <- genus_table %>%
  filter(rowSums(across(everything())) >= 0) 
dim(genus_table_filtered)

# Total number of reads per sample
summary(colSums(genus_table))

# Total reads per family
summary(rowSums(genus_table))


taxa_table_genus_filtered <- taxa_table_genus[rownames(taxa_table_genus) %in% rownames(genus_table_filtered), ]
dim(taxa_table_genus_filtered)


# Make phyloseq object - raw reads
otu_table_phyloseq <- otu_table(as.matrix(genus_table), taxa_are_rows = TRUE)
sample_data_phyloseq <- sample_data(metadata)
taxonomy_phyloseq <- tax_table(as.matrix(taxa_table_genus))

ps_rel <- phyloseq(otu_table_phyloseq, sample_data_phyloseq, taxonomy_phyloseq)
ps_rel

# saveRDS(ps_rel, file = "phyloseq_object_genus_rel.rds")

```

### Session Info
```{r}
sessionInfo()
```