---
title: "The assembly based on Sloan's Neutral Model"
author: "Jia Xiu (x.jia@rug.nl)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
html_preview: false
editor_options: 
  chunk_output_type: console
---

refer: https://github.otu_table/seb369/landuse_otu_tablem_assembly
Barnett SE, Youngblut ND, Buckley DH. Soil characteristics and land-use drive bacterial community assembly patterns, FEMS Microbiology Ecology, 2020;96(1):fiz194, doi: 10.1093/femsec/fiz194
Burns, A., Stephens, W., Stagaman, K. et al. Contribution of neutral processes to the assembly of gut microbial communities in the zebrafish over host development. ISME J 10, 655-664 (2016) doi:10.1038/ismej.2015.142
https://github.otu_table/seb369/landuse_otu_tablem_assembly/blob/master/Neutral_model.md
https://static-content.springer.otu_table/esm/art%3A10.1038%2Fismej.2015.142/MediaObjects/41396_2016_BFismej2015142_MOESM54_ESM.txt


### Initiate libraries
```{r, message=FALSE}
# Clear workspace
rm(list=ls())

# load libraries

library(phyloseq) 
library(tidyverse)
# library(tibble)
library(RColorBrewer)
library(minpack.lm)
library(Hmisc)
library(ggpubr) 

# plot theme
mytheme <- theme_light()+
  theme(text = element_text(size = 10),
        plot.title = element_text(size = 10),
        legend.title = element_text(face = "bold", size = 10),
        legend.text = element_text(size = 10), 
        legend.justification = c("right", "top"),
        legend.background = element_rect(),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 10, color = "black"),
        axis.ticks = element_line(colour = "grey70", linewidth = 0.2),
        panel.grid.major = element_line(colour = "grey70", linewidth = 0.1),
        panel.grid.minor = element_blank())  

# Path to the folder containing Kraken2 report files
folder_path <- "sloan_neutral_model_species/"

# Create the directory if it does not already exist
if (!dir.exists(folder_path)) {
  dir.create(folder_path)
}

```


## 1. load dataset for analysis
```{r}
# read phyloseq object from all the samples - raw reads
ps_rel <- readRDS("phyloseq_object_species_rel.rds")

# only keep the samples from microcosm experiment
ps_microcosm <- subset_samples(ps_rel, category == "microcosm")
# remove taxa do not appear in any samples
ps_microcosm <- filter_taxa(ps_microcosm, function(x) sum(x) != 0, TRUE)

otu_table <- as.data.frame(otu_table(ps_microcosm)) 

otu_table <- t(otu_table)
dim(otu_table)
otu_table[1:5, 1:2]
range(rowSums(otu_table))

# classification based on ALDEx2 analysis
taxa_classified <- read.csv("/Users/xiujia/Library/CloudStorage/OneDrive-TechnischeUniversitätIlmenau/Data_analysis/01_Mixing_exp._04-2023/01_16S_amplicon/01_emu_results/ALDEx2_differential_analysis_source_coms_species/ALDEx2_results_all_passage.csv", header = TRUE, row.names = 1) %>% 
  select(-c(we.ep, we.eBH, rab.all, rab.win.b, rab.win.f, diff.btw, diff.win, effect, overlap, wi.ep, wi.eBH, significant, text_label))

head(taxa_classified)

```

## 2 Sloan's neutral model for the total community
```{r}
# Calculate the number of individuals per community (based on relative abundance OTU table)
range(apply(otu_table, 1, sum))

# Calculate the average relative abundance of each taxa across communities
p.m <- apply(otu_table, 2, mean)
p.m <- p.m / 100
p.m <- p.m[p.m != 0]
p.df <- data.frame(p.m) %>% rownames_to_column(var="taxon")
str(p.df)
head(p.df)

# Calculate the occurrence frequency of each taxa
otu_table.bi <- 1*(otu_table>0)
freq <- apply(otu_table.bi, 2, mean)
freq <- freq[freq != 0]
freq.df <- data.frame(freq) %>% rownames_to_column(var="taxon")
str(freq.df)
head(freq.df)

# otu_tablebine the average relative abundance and occurrence of each taxa
df <- inner_join(p.df, freq.df, by="taxon") %>% arrange(p.m)

# Remove rows with any zero (absent in either source pool or local communities). double check!
df <- df %>% filter(freq != 0, p.m != 0)
str(df)
p <- df$p.m
freq <- df$freq

                             
#Calculate the limit of detection
d <- 0.001

##Fit model parameter m (or Nm) using Non-linear least squares (NLS)
# N is the number of individuals/reads in a community, which you calculated as 100
N <- 100
m.fit <- nlsLM(freq ~ pbeta(d, N*m*p, N*m*(1-p), lower.tail=FALSE), start=list(m=0.1))
m.ci <- confint(m.fit, 'm', level=0.95)
m.sum <- summary(m.fit)
m.coef <- coef(m.fit)

##Calculate goodness-of-fit (R-squared and Root Mean Squared Error)
freq.pred <- pbeta(d, N*coef(m.fit)*p, N*coef(m.fit)*(1-p), lower.tail=FALSE)
Rsqr <- 1 - (sum((freq - freq.pred)^2))/(sum((freq - mean(freq))^2))


# Get table of model fit stats
fitstats <- data.frame(m=m.coef, m.low.ci=m.ci[1], m.up.ci=m.ci[2], 
                       Rsqr=Rsqr, p.value=m.sum$parameters[4], N=N, 
                       Samples=nrow(otu_table), Richness=length(p), 
                       Detect=d)

# Get confidence interval for predictions
freq.pred.ci <- binconf(freq.pred*nrow(otu_table), nrow(otu_table), alpha=0.05, method="wilson", return.df=TRUE)

# Get table of predictions
pred.df <- data.frame(metacom_RA=p, frequency=freq.pred, 
                      frequency_lowerCI=freq.pred.ci[,2], 
                      frequency_upperCI=freq.pred.ci[,3]) %>%
  unique()

head(pred.df)

# Get table of observed occupancy and abundance
obs.df <- df %>%
  rename(metacom_RA = p.m, frequency=freq)

obs.df <- left_join(obs.df, taxa_classified , by = c('taxon')) %>%
  mutate(enrichment = ifelse(is.na(enrichment), "New emergence", enrichment)) %>% 
  column_to_rownames('taxon') %>% 
  mutate(enrichment = factor(enrichment, levels = c("Freshwater", "New emergence", "Not significant", "Seawater"))) %>% mutate(classification = ifelse(enrichment == "New emergence", "No", "Yes"))

head(obs.df)

table(obs.df$enrichment)

# scatter plot with fit lines for all sub-communities
(p_all <- ggplot(data=obs.df) +
    geom_line(data=pred.df, aes(x=log10(metacom_RA), y=frequency), color="black", alpha=.7) + 
    geom_line(data=pred.df, aes(x=log10(metacom_RA), y=frequency_lowerCI), linetype=2, color="red", alpha=.7) + 
    geom_line(data=pred.df, aes(x=log10(metacom_RA), y=frequency_upperCI), linetype=2, color="red", alpha=.7) + 
    geom_point(data=obs.df, aes(x=log10(metacom_RA), y=frequency, color=enrichment, shape = classification), alpha=.7, size=1) +
    geom_text(data=fitstats, aes(label = paste("R^2 == ", round(Rsqr, 3))), 
              x=-5.2, y=0.9, size=4, parse=TRUE) +
    geom_text(data=fitstats, aes(label = paste("italic(m) ==", round(m, 3))), 
              x=-5.2, y=0.8, size=4, parse=TRUE) + 
    scale_x_continuous(limits = c(-6.3, -1)) + 
    scale_y_continuous(limits = c(0, 1)) + 
    labs(x="Log10 abundance in\nmetacommunity", y="Frequency detected", title = "All sub-communities") +
    scale_color_manual(values = c("Freshwater" = "#0073C2FF", 
                                  "Seawater" = "#CD534CFF", 
                                  "Not significant" = "darkgray")) +
    scale_shape_manual(values=c(1, 19)) +
    # scale_colour_manual(values=my_palette) + 
    guides(colour = guide_legend(title="Habitat preference"), shape = guide_legend(title = "Deteched in \nsource communities"))+
    mytheme
)

ggsave(paste0(folder_path, "Sloan_neutral_model_whole_microcosm.png"), p_all, width = 3.2, height = 3.2)

```


## 3. Sloan's neutral model for communities at freshwater
```{r}

otu_table_fresh <- otu_table[grep("_f_", rownames(otu_table)), ]
otu_table_fresh[1:2, 1:2]

# Calculate the number of individuals per community (based on relative abundance OTU table)
range(apply(otu_table_fresh, 1, sum))

# Calculate the average relative abundance of each taxa across communities
p.m <- apply(otu_table_fresh, 2, mean)
p.m <- p.m / 100
p.m <- p.m[p.m != 0]
p.df <- data.frame(p.m) %>% rownames_to_column(var="taxon")
str(p.df)
head(p.df)

# Calculate the occurrence frequency of each taxa
otu_table_fresh.bi <- 1*(otu_table_fresh > 0)
freq <- apply(otu_table_fresh.bi, 2, mean)
freq <- freq[freq != 0]
freq.df <- data.frame(freq) %>% rownames_to_column(var="taxon")
str(freq.df)
head(freq.df)

# combine the average relative abundance and occurrence of each taxa
df <- inner_join(p.df, freq.df, by="taxon") %>% arrange(p.m)

# Remove rows with any zero (absent in either source pool or local communities). double check!
df <- df %>% filter(freq != 0, p.m != 0)
str(df)

p <- df$p.m
freq <- df$freq

                                
#Calculate the limit of detection
d <- 0.001

##Fit model parameter m (or Nm) using Non-linear least squares (NLS)
# N is the number of individuals/reads in a community, which you calculated as 100
N <- 100
m.fit <- nlsLM(freq ~ pbeta(d, N*m*p, N*m*(1-p), lower.tail=FALSE), start=list(m=0.1))
m.ci <- confint(m.fit, 'm', level=0.95)
m.sum <- summary(m.fit)
m.coef <- coef(m.fit)

##Calculate goodness-of-fit (R-squared and Root Mean Squared Error)
freq.pred <- pbeta(d, N*coef(m.fit)*p, N*coef(m.fit)*(1-p), lower.tail=FALSE)
Rsqr <- 1 - (sum((freq - freq.pred)^2))/(sum((freq - mean(freq))^2))

# Get table of model fit stats
fitstats <- data.frame(m=m.coef, m.low.ci=m.ci[1], m.up.ci=m.ci[2], 
                       Rsqr=Rsqr, p.value=m.sum$parameters[4], N=N, 
                       Samples=nrow(otu_table_fresh), Richness=length(p), 
                       Detect=d)

# Get confidence interval for predictions
freq.pred.ci <- binconf(freq.pred*nrow(otu_table_fresh), nrow(otu_table_fresh), alpha=0.05, method="wilson", return.df=TRUE)

# Get table of predictions
pred.df <- data.frame(metacom_RA=p, frequency=freq.pred, 
                      frequency_lowerCI=freq.pred.ci[,2], 
                      frequency_upperCI=freq.pred.ci[,3]) %>%
  unique()

head(pred.df)


# Get table of observed occupancy and abundance
obs.df <- df %>%
  rename(metacom_RA = p.m, frequency=freq)

obs.df <- left_join(obs.df, taxa_classified , by = c('taxon')) %>%
  mutate(enrichment = ifelse(is.na(enrichment), "New emergence", enrichment)) %>% 
  column_to_rownames('taxon') %>% 
  mutate(enrichment = factor(enrichment, levels = c("Freshwater", "New emergence", "Not significant", "Seawater"))) %>% mutate(classification = ifelse(enrichment == "New emergence", "No", "Yes"))

head(obs.df)

table(obs.df$enrichment)

(p_freshwater <- ggplot(data=obs.df) +
    geom_line(data=pred.df, aes(x=log10(metacom_RA), y=frequency), color="black", alpha=.7) + 
    geom_line(data=pred.df, aes(x=log10(metacom_RA), y=frequency_lowerCI), linetype=2, color="red", alpha=.7) + 
    geom_line(data=pred.df, aes(x=log10(metacom_RA), y=frequency_upperCI), linetype=2, color="red", alpha=.7) + 
    geom_point(data=obs.df, aes(x=log10(metacom_RA), y=frequency, color=enrichment, shape = classification), alpha=.7, size=1) +
    geom_text(data=fitstats, aes(label = paste("R^2 == ", round(Rsqr, 3))), 
              x=-5.2, y=0.9, size=4, parse=TRUE) +
    geom_text(data=fitstats, aes(label = paste("italic(m) ==", round(m, 3))), 
              x=-5.2, y=0.8, size=4, parse=TRUE) + 
    scale_x_continuous(limits = c(-6.3, -1)) + 
    scale_y_continuous(limits = c(0, 1)) + 
    labs(x="Log10 abundance in\nmetacommunity", y="Frequency detected", title = "Freshwater") +
    scale_color_manual(values = c("Freshwater" = "#0073C2FF", 
                                  "Seawater" = "#CD534CFF", 
                                  "Not significant" = "darkgray")) +
    scale_shape_manual(values=c(1, 19)) +
    # scale_colour_manual(values=my_palette) + 
    guides(colour = guide_legend(title="Habitat preference"), shape = guide_legend(title = "Deteched in \nsource communities"))+
    mytheme
)

ggsave(paste0(folder_path, "Sloan_neutral_model_freswater19.png"), p_fresh, width = 5, height = 3.2)

```

## 4. Sloan's neutral model for communities at seawater
```{r}

otu_table_sea <- otu_table[grep("_b_", rownames(otu_table)), ]
otu_table_sea[1:2, 1:2]

# Calculate the number of individuals per community (based on relative abundance OTU table)
range(apply(otu_table_sea, 1, sum))

# Calculate the average relative abundance of each taxa across communities
p.m <- apply(otu_table_sea, 2, mean)
p.m <- p.m / 100
p.m <- p.m[p.m != 0]
p.df <- data.frame(p.m) %>% rownames_to_column(var="taxon")
str(p.df)
head(p.df)

# Calculate the occurrence frequency of each taxa
otu_table_sea.bi <- 1*(otu_table_sea>0)
freq <- apply(otu_table_sea.bi, 2, mean)
freq <- freq[freq != 0]
freq.df <- data.frame(freq) %>% rownames_to_column(var="taxon")
str(freq.df)
head(freq.df)

# combine the average relative abundance and occurrence of each taxa
df <- inner_join(p.df, freq.df, by="taxon") %>% arrange(p.m)

# Remove rows with any zero (absent in either source pool or local communities). double check!
df <- df %>% filter(freq != 0, p.m != 0)
str(df)

p <- df$p.m
freq <- df$freq

                                
#Calculate the limit of detection
d <- 0.001

##Fit model parameter m (or Nm) using Non-linear least squares (NLS)
# N is the number of individuals/reads in a community, which you calculated as 100
N <- 100
m.fit <- nlsLM(freq ~ pbeta(d, N*m*p, N*m*(1-p), lower.tail=FALSE), start=list(m=0.1))
m.ci <- confint(m.fit, 'm', level=0.95)
m.sum <- summary(m.fit)
m.coef <- coef(m.fit)

##Calculate goodness-of-fit (R-squared and Root Mean Squared Error)
freq.pred <- pbeta(d, N*coef(m.fit)*p, N*coef(m.fit)*(1-p), lower.tail=FALSE)
Rsqr <- 1 - (sum((freq - freq.pred)^2))/(sum((freq - mean(freq))^2))


# Get table of model fit stats
fitstats <- data.frame(m=m.coef, m.low.ci=m.ci[1], m.up.ci=m.ci[2], 
                       Rsqr=Rsqr, p.value=m.sum$parameters[4], N=N, 
                       Samples=nrow(otu_table_sea), Richness=length(p), 
                       Detect=d)

# Get confidence interval for predictions
freq.pred.ci <- binconf(freq.pred*nrow(otu_table_sea), nrow(otu_table_sea), alpha=0.05, method="wilson", return.df=TRUE)

# Get table of predictions
pred.df <- data.frame(metacom_RA=p, frequency=freq.pred, 
                      frequency_lowerCI=freq.pred.ci[,2], 
                      frequency_upperCI=freq.pred.ci[,3]) %>%
  unique()

head(pred.df)


# Get table of observed occupancy and abundance
obs.df <- df %>%
  rename(metacom_RA = p.m, frequency=freq)

obs.df <- left_join(obs.df, taxa_classified , by = c('taxon')) %>%
  mutate(enrichment = ifelse(is.na(enrichment), "New emergence", enrichment)) %>% 
  column_to_rownames('taxon') %>% 
  mutate(enrichment = factor(enrichment, levels = c("Freshwater", "New emergence", "Not significant", "Seawater"))) %>% mutate(classification = ifelse(enrichment == "New emergence", "No", "Yes"))

head(obs.df)

table(obs.df$enrichment)

(p_seawater <- ggplot(data=obs.df) +
    geom_line(data=pred.df, aes(x=log10(metacom_RA), y=frequency), color="black", alpha=.7) + 
    geom_line(data=pred.df, aes(x=log10(metacom_RA), y=frequency_lowerCI), linetype=2, color="red", alpha=.7) + 
    geom_line(data=pred.df, aes(x=log10(metacom_RA), y=frequency_upperCI), linetype=2, color="red", alpha=.7) + 
    geom_point(data=obs.df, aes(x=log10(metacom_RA), y=frequency, color=enrichment, shape = classification), alpha=.7, size=1) +
    geom_text(data=fitstats, aes(label = paste("R^2 == ", round(Rsqr, 3))), 
              x=-5.2, y=0.9, size=4, parse=TRUE) +
    geom_text(data=fitstats, aes(label = paste("italic(m) ==", round(m, 3))), 
              x=-5.2, y=0.8, size=4, parse=TRUE) + 
    scale_x_continuous(limits = c(-6.3, -1)) + 
    scale_y_continuous(limits = c(0, 1)) + 
    labs(x="Log10 abundance in\nmetacommunity", y="Frequency detected", title = "Seawater") +
    scale_color_manual(values = c("Freshwater" = "#0073C2FF", 
                                  "Seawater" = "#CD534CFF", 
                                  "Not significant" = "darkgray")) +
    scale_shape_manual(values=c(1, 19)) +
    # scale_colour_manual(values=my_palette) + 
    guides(colour = guide_legend(title="Habitat preference"), shape = guide_legend(title = "Deteched in \nsource communities"))+
    mytheme)


ggsave(paste0(folder_path, "Sloan_neutral_model_seawater.png"), p_fresh, width = 5, height = 3.2)

```

```{r}

# combine all the plots

(p <- ggarrange(p_all, p_freshwater, p_seawater, labels = c("A", "B", "C"), 
  common.legend = TRUE, legend = "right", ncol = 1, nrow = 3, align = "hv"))

# ggsave(paste0(folder_path, "Fig_5_Sloan_neutral_model_all_freshwater_seawater.png"), p, width = 4, height = 7.5)
# ggsave(paste0(folder_path, "Fig_5_Sloan_neutral_model_all_freshwater_seawater_raw.pdf"), p, width = 4, height = 7.5)

# only freshwater and seawater
(p <- ggarrange(p_freshwater, p_seawater, labels = c("A", "B"), 
  common.legend = TRUE, legend = "right", ncol = 2, nrow = 1, align = "hv"))

# ggsave(paste0(folder_path, "Fig_5_Sloan_neutral_model_freshwater_seawater.png"), p, width = 6.5, height = 2.5)
# ggsave(paste0(folder_path, "Fig_5_Sloan_neutral_model_freshwater_seawater_raw.pdf"), p, width = 6.5, height = 2.5)

```

### Session Info
```{r}
sessionInfo()
```